---
title: 'Case Study 4: Association of Brain Function and Structure with Neuropsychological
  Measures'
author: "Emmanuel Fle Chea"
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
# Load needed libraries
library(dplyr)
library(tidyverse)    
library(readr)        
library(broom)
library(knitr)
library(lm.beta)
```


```{r, message=FALSE, warning=FALSE}
dti_data <- read.csv("C:/Users/emman/OneDrive/Desktop/MPH Data Science - UMN/Biostatistical Consulting/Case_Study4/DTI data.csv", stringsAsFactors = FALSE)

# Inspect the structure and summary of the dataset
str(dti_data)
view(dti_data)
colnames(dti_data)
```

```{r}
# Check the raw values and data type
str(dti_data$Group)
```


```{r}
# 2. Check raw Group data
str(dti_data$Group)
unique(dti_data$Group)
```


```{r}
# (4) Verify the result
str(dti_data$Group)
unique(dti_data$Group)
table(dti_data$Group)
```





```{r}
# Rename unusual column names to simpler, clean versions
names(dti_data)[which(names(dti_data) == "AGE..mo.")] <- "Age_mo"
names(dti_data)[which(names(dti_data) == "Medical.Risk.Factors")] <- "MedRisk"

# Recode MedRisk to binary: 1 if > 0, else 0
dti_data$MedRiskBinary <- ifelse(dti_data$MedRisk > 0, 1, 0)

# Create age in years
dti_data$Age_yrs <- dti_data$Age_mo / 12

# Convert Group to factor
dti_data$Group <- factor(
  dti_data$Group,
  levels = c(1, 2),
  labels = c("attenuated", "severe")
)

# Create summary table (Table 1) including both % Female and % Male
descriptives_table1 <- dti_data %>%
  group_by(Group) %>%
  summarise(
    n = n(),
    age_mean = round(mean(Age_yrs, na.rm = TRUE), 2),
    age_sd = round(sd(Age_yrs, na.rm = TRUE), 2),
    percent_female = round(mean(Sex == "female", na.rm = TRUE) * 100, 1),
    percent_male = round(mean(Sex == "male", na.rm = TRUE) * 100, 1),
    med_risk_pct = round(mean(MedRiskBinary, na.rm = TRUE) * 100, 1),
    FSIQ_mean = round(mean(FSIQ, na.rm = TRUE), 2),
    FSIQ_sd = round(sd(FSIQ, na.rm = TRUE), 2),
    CVLT_mean = round(mean(CVLT, na.rm = TRUE), 2),
    CVLT_sd = round(sd(CVLT, na.rm = TRUE), 2),
    JLO_mean = round(mean(JLO, na.rm = TRUE), 2),
    JLO_sd = round(sd(JLO, na.rm = TRUE), 2),
    FA_mean = round(mean(FA, na.rm = TRUE), 3),
    FA_sd = round(sd(FA, na.rm = TRUE), 3),
    MD_mean = round(mean(MD, na.rm = TRUE), 3),
    MD_sd = round(sd(MD, na.rm = TRUE), 3),
    TOVAd_mean = round(mean(TOVAd, na.rm = TRUE), 2),
    TOVAd_sd = round(sd(TOVAd, na.rm = TRUE), 2),
    TOVAOm_mean = round(mean(TOVAOm, na.rm = TRUE), 2),
    TOVAOm_sd = round(sd(TOVAOm, na.rm = TRUE), 2),
    TOVACom_mean = round(mean(TOVACom, na.rm = TRUE), 2),
    TOVACom_sd = round(sd(TOVACom, na.rm = TRUE), 2),
    TOVART_mean = round(mean(TOVART, na.rm = TRUE), 2),
    TOVART_sd = round(sd(TOVART, na.rm = TRUE), 2),
    TovaVAR_mean = round(mean(TovaVAR, na.rm = TRUE), 2),
    TovaVAR_sd = round(sd(TovaVAR, na.rm = TRUE), 2)
  )

# Display Table 1
kable(descriptives_table1, caption = "Table 1: Demographics, Medical Risk, and Neurocognitive Measures by Patient Group")
```

```{r}
colnames(dti_data)
```


```{r}
dti_data$MedRiskBinary
```



```{r}
# Compare groups on DTI measures using t-tests
t_test_FA  <- t.test(FA ~ Group, data = dti_data)
t_test_MD  <- t.test(MD ~ Group, data = dti_data)

cat("T-test for FA by Group\n")
print(t_test_FA)

cat("\nT-test for MD by Group\n")
print(t_test_MD)
```
# Correlation Analysis within Groups

```{r}
# Filter the data for each group
severe_data <- filter(dti_data, Group == "severe")
attenuated_data <- filter(dti_data, Group == "attenuated")
```

# Calculate correlation matrix for severe group (for FA, MD, and TOVAd)

```{r}
# Calculate correlation matrix for severe group (for FA, MD, and TOVAd)
cor_severe <- cor(severe_data[c("FA", "MD", "TOVAd")], use = "complete.obs")

# Calculate correlation matrix for attenuated group (for FA, MD, and TOVAd)
cor_attenuated <- cor(attenuated_data[c("FA", "MD", "TOVAd")], use = "complete.obs")

# Print the correlation matrices
cat("Correlation Matrix for Severe Group:\n")
print(cor_severe)
cat("\nCorrelation Matrix for Attenuated Group:\n")
print(cor_attenuated)
```
# Regression Models to Explore Association and Interactions

```{r}
# The study hypothesis suggests that the associations between TOVA measures and DTI measures (FA and MD)
# might differ between groups. One way to test this is to include an interaction term in regression models.
# We will demonstrate with the TOVA d-prime measure.
#
# Model 1: TOVA_dprime predicted by FA, Group, and their interaction.
model_FA <- lm(TOVAd ~ FA + Group + FA * Group, data = dti_data)
summary(model_FA)
```


```{r}
# Model 2: TOVA_dprime predicted by MD, Group, and their interaction.
model_MD <- lm(TOVAd ~ MD + Group + MD * Group, data = dti_data)
summary(model_MD)
```


```{r}
# Model 3: Adjust for both DTI measures simultaneously:
model_both <- lm(TOVAd ~ FA + MD + Group + FA * Group + MD * Group, data = dti_data)
summary(model_both)
```


```{r}
# Loop Over TOVA Measures
tova_measures <- c("TOVAd", "TOVAOm", "TOVACom", "TOVART", "TovaVAR")

# Loop: For each TOVA measure, predict it from FA, MD, Group, and their interactions.
results_list <- list()
for (measure in tova_measures) {
  # Build the regression formula dynamically: Outcome ~ FA + MD + Group + FA * Group + MD * Group
  formula_str <- paste(measure, "~ FA + MD + Group + FA * Group + MD * Group")
  model <- lm(as.formula(formula_str), data = dti_data)
  results_list[[measure]] <- summary(model)
  
  cat("\n---------------------------------------------------\n")
  cat("Regression results for", measure, ":\n")
  print(summary(model))
}

# Output
tidy_model_FA <- broom::tidy(model_FA)
kable(tidy_model_FA)
```

```{r, message=FALSE, warning=FALSE}
# Visualization of Key Relationships
# Visualize the relationship between FA and TOVAd by Group.
library(ggplot2)
ggplot(dti_data, aes(x = FA, y = TOVAd, color = Group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Association between FA and TOVAd by Group",
       x = "Fractional Anisotropy (FA)",
       y = "TOVAd")

# Similarly, visualize the relationship between MD and TOVAd.
ggplot(dti_data, aes(x = MD, y = TOVAd, color = Group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Association between MD and TOVAd by Group",
       x = "Mean Diffusivity (MD)",
       y = "TOVAd")
```

# Diagnostic plots for the final model with both FA and MD

```{r}
par(mfrow = c(2, 2))
plot(model_both)  
par(mfrow = c(1, 1))
```

# R-squared and Adjusted R-squared

```{r}
# Extract R-squared values
model_summaries <- list(model_FA, model_MD, model_both) %>%
  lapply(summary)

r2_table <- data.frame(
  Model = c("FA Only", "MD Only", "FA + MD + Group + Interactions"),
  R_Squared = sapply(model_summaries, function(x) round(x$r.squared, 3)),
  Adj_R_Squared = sapply(model_summaries, function(x) round(x$adj.r.squared, 3))
)

kable(r2_table, caption = "Table: R-squared and Adjusted R-squared for Models")
```

# AIC for Model Fit Comparison

```{r}
aic_table <- data.frame(
  Model = c("FA Only", "MD Only", "Both + Interactions"),
  AIC = c(AIC(model_FA), AIC(model_MD), AIC(model_both))
)

kable(aic_table, caption = "Table: AIC Model Fit Comparison")
```

# Sensitivity Analysis: Including MedRisk

```{r}
model_medrisk <- lm(TOVAd ~ FA + MD + Group + FA * Group + MD * Group + MedRiskBinary, data = dti_data)
summary(model_medrisk)

cat("Adjusted R-squared with MedRisk:\n")
summary(model_medrisk)$adj.r.squared
```

# Sensitivity Analysis: Exclude Influential Outliers

```{r}
# Cook's distance for identifying outliers
cooks_d <- cooks.distance(model_both)
threshold <- 4 / nrow(dti_data)

# Exclude influential points
dti_data_cleaned <- dti_data[cooks_d < threshold, ]

# Refit model
model_cleaned <- lm(TOVAd ~ FA + MD + Group + FA * Group + MD * Group, data = dti_data_cleaned)
summary(model_cleaned)

cat("Adjusted R-squared after removing influential points:\n")
summary(model_cleaned)$adj.r.squared
```

# Strength of Association: Standardized Coefficients

```{r}
# Standardized regression coefficients for effect size comparison
model_both_std <- lm.beta(model_both)

# Tidy output
tidy_std <- broom::tidy(model_both_std)
kable(tidy_std, caption = "Table: Standardized Coefficients (Effect Sizes) for Model 3")
```

